---
title: "Nursing Home Case Study"
author: "Zhouming Sun"
date: "2025-02-08"
output:
  word_document: default
  html_document: default
  pdf_document: default
header-includes: \usepackage{fvextra} \DefineVerbatimEnvironment{Highlighting}{Verbatim}{breaklines,commandchars=\\\{\}}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(usmap)
```

## Correlations and Plots

First, a quick look into the *NH_ProviderInfo_Nov2024.csv* file reveals correlations about the providers and various measures of their quality. The easiest measure to compare against is each provider's Overall Rating, on a scale of 1-5, with 1 being the lowest score. 

```{r providers turnover}
providers <- read.csv(r"(C:\Users\pokem\Downloads\nursing_homes_including_rehab_services_current_data\NH_ProviderInfo_Nov2024.csv)")

aggregate(Total.nursing.staff.turnover ~ Overall.Rating, data=providers, FUN = mean) %>% plot(main = "Overall Rating ~ Mean Total Nursing Staff Turnover")

```

This demonstrates that high nursing staff turnover is correlated with a poor overall rating. If one is focused on contractor staffing data, which is not available in the *NH_ProviderInfo_Nov2024.csv* file, total nursing staff turnover will have to do as a proxy measurement. 

The following graphs show several other correlations. 

```{r providers other aggregations}

aggregate(Number.of.administrators.who.have.left.the.nursing.home ~ Overall.Rating, data=providers, FUN = mean) %>% plot(main = "Overall Rating ~ Mean Number of departed administrators")

aggregate(Number.of.Facility.Reported.Incidents ~ Overall.Rating, data=providers, FUN = mean) %>% plot(main = "Overall Rating ~ Mean Number of Facility Reported Incidents")

aggregate(Number.of.Substantiated.Complaints ~ Overall.Rating, data=providers, FUN = mean) %>% plot(main = "Overall Rating ~ Mean Number of Substantiated Complaints")

aggregate(Total.Number.of.Penalties ~ Overall.Rating, data=providers, FUN = mean) %>% plot(main = "Overall Rating ~ Mean Number of Penalties")

aggregate(Number.of.Payment.Denials ~ Overall.Rating, data=providers, FUN = sum) %>% plot(main = "Overall Rating ~ Total Number of Payment Denials", sub="Note: Payment denials are from Medicaid or Medicare" )

aggregate(Reported.Total.Nurse.Staffing.Hours.per.Resident.per.Day ~ Overall.Rating, data=providers, FUN = mean) %>% plot(
  main = "Overall Rating ~ Nurse Hours Per Resident",
  ylab = "Daily Mean Nurse Hours per Resident", 
  yaxp = c(3,5,8)
)
```

From these graphs, it is clear that Overall Rating is negatively correlated with 'general indicators of poor quality', being in this case: staff turnover, administrator turnover, reported facility incidents, substantiated complaints, penalties, and payment denials from Medicaid or Medicare. 

So it might be a good idea to avoid association with these poorly rated facilities. Alternatively, they could be prime customers for contractors. However, it would be in bad taste to simply churn a stream of nurse contractors through terribly managed facilities. 

The last graph demonstrates a correlation between mean nurse hours per resident and overall rating. Facilities that are struggling with low rating and quality would be well advised add more staff such that the Daily Mean Nurse Hours are above 4 per resident.  

## Facility Ownership

Being aware of poorly managed and rated facilities is nice, but the *NH_Ownership_Nov2024.csv* file can be utilized to identify the owners responsible for the penalties. The following code connects each facility's penalties to the owners at the time. This is to not count cases for owners if their association date comes after the penalty date. In other words, if you just joined a facility, you are not responsible for the facility's past penalties. If a penalty occurs after you become an owner, it is counted against you. 

```{r ownership}
ownership <- read.csv(r"(C:\Users\pokem\Downloads\nursing_homes_including_rehab_services_current_data\NH_Ownership_Nov2024.csv)")
penalties <- read.csv(r"(C:\Users\pokem\Downloads\nursing_homes_including_rehab_services_current_data\NH_Penalties_Nov2024.csv)")

ownership$DATE <- sapply(strsplit(ownership$Association.Date, split = " ", fixed = TRUE), function(x)(x[2])) %>% as.Date("%m/%d/%Y")

for (i in 1:nrow(penalties)) {
  owners <- ownership[which(ownership$DATE < penalties$Penalty.Date[i] & 
                              ownership$CMS.Certification.Number..CCN. == penalties$CMS.Certification.Number..CCN.[i]) , 
                      c("Owner.Name", "CMS.Certification.Number..CCN.", "DATE")]
  
  owners <- owners[! duplicated(owners$Owner.Name), ] # remove duplicate owners for a single penalty
  
  if (i == 1) {
    penalized_owners_with_date <- owners
  } else{
    penalized_owners_with_date <- bind_rows(penalized_owners_with_date, owners)
  }
}

count_owner_penalty_with_date <- count(penalized_owners_with_date, Owner.Name)

summary(count_owner_penalty_with_date)
```

This summary shows most penalized owners keep it to 5 or fewer penalties. Then there is a special subset of the bunch who are responsible for the vast majority of penalties. Let's take a look at the top 50 offenders. 

```{r}
head(count_owner_penalty_with_date[order(-count_owner_penalty_with_date$n), ] , 50)
```

We can use this list as a blacklist of sorts. It would be advisable to avoid working with facilities that are associated with any of these owners, as they have a history of collecting penalties, which is negatively correlated with Overall Rating and Staff Turnover. 

## Maps using FIPS Codes

By joining the `providers` file with the `daily_nurse_staffing` dataframe, we can combine the county and state FIPS codes for mapping purposes. The inner join conveniently selects only for providers with data in the 2024 Q2 timeframe of `daily_nurse_staffing`. 

```{r}
daily_nurse_staffing <- read.csv(r"(C:\Users\pokem\Downloads\PBJ_Daily_Nurse_Staffing_Q2_2024\PBJ_Daily_Nurse_Staffing_Q2_2024.csv)")

providers <- inner_join(providers, daily_nurse_staffing[!duplicated(daily_nurse_staffing$PROVNUM), c("PROVNUM", "COUNTY_FIPS")], by = c("CMS.Certification.Number..CCN." = "PROVNUM"))
providers <- relocate(providers, COUNTY_FIPS, .after = State)

providers$COUNTY_FIPS <- as.character(providers$COUNTY_FIPS)
providers$COUNTY_FIPS <- str_pad(providers$COUNTY_FIPS, 3, pad = "0") # county FIPS codes should be 3 digits long, and can begin with a 0

providers$STATE_FIPS <- usmap::fips(state = providers$State)
providers <- relocate(providers, STATE_FIPS, .after = State)

providers$fips <- paste0(providers$STATE_FIPS, providers$COUNTY_FIPS)
providers <- relocate(providers, fips, .after = State)


df_rating <- aggregate(Overall.Rating ~ fips, data = providers, FUN = mean)
plot_usmap(
  color = "white", 
  linewidth = 0.1, 
  regions = "counties", 
  data = df_rating, 
  values = "Overall.Rating"
) + 
  scale_fill_gradient(
    labels = scales::label_number(accuracy = 1), 
    high = "green", 
    low = 'red'
  ) + 
  theme(
    legend.position = 'top'  , 
    text = element_text(size = 10)
  ) +
  labs(fill = "Overall Rating") + 
  guides(fill = guide_colorbar(
    barwidth = unit(10, 'cm')
  )
  )

```

It would be worth an investigation into why some counties are associated with lower rated facilities, and vice versa. 


```{r echo=FALSE}
df_turnover <- aggregate(Total.nursing.staff.turnover ~ fips, data = providers, FUN = mean)
plot_usmap(
  color = "white", 
  linewidth = 0.1, 
  regions = "counties", 
  data = df_turnover, 
  values = "Total.nursing.staff.turnover"
) + 
  scale_fill_gradient(
    labels = scales::label_number(accuracy = 1), 
    high = "red", 
    low = 'white'
  ) + 
  theme(
    legend.position = 'top'  , 
    text = element_text(size = 10)
  ) +
  labs(fill = "Total Nursing Staff Turnover") + 
  guides(fill = guide_colorbar(
    barwidth = unit(10, 'cm')
  )
  )


df_penalty_mean <- aggregate(Total.Number.of.Penalties ~ fips, data = providers, FUN = mean)
plot_usmap(
  color = "white", 
  linewidth = 0.1, 
  regions = "counties", 
  data = df_penalty_mean, 
  values = "Total.Number.of.Penalties"
) + 
  scale_fill_gradient(
    trans = 'log',
    labels = scales::label_number(accuracy = 1), 
    high = "red", 
    low = 'white'
  ) + 
  theme(
    legend.position = 'top'  , 
    text = element_text(size = 10)
  ) +
  labs(fill = "Mean Number of Penalties") + 
  guides(fill = guide_colorbar(
    barwidth = unit(10, 'cm')
  )
  )

```

Further investigation into counties with high Total Nursing Staff Turnover and Mean Number of Penalties might be fruitful. 
These maps demonstrate that there are hotspot regions, and the problematic facilities are not evenly spread throughout the United States. 

## Conclusions

If possible, avoid doing business known penalty-collecting owners and facilities. If that is unavoidable, an entity should insist upon changes (that correlate with a better overall rating) before allowing them to match with its contractors. Adding enough nurses/staff such that the mean nurse staffing hours per resident per day is greater than 4 would be ideal. Perhaps requiring facilities to sign on multiple contractors at once could be a solution to the low rating and high nurse turnover problem. 
